\pkg{cleanR}: Maid for Cleaning Datasets in \proglang{R}

### 2016-12-29 latex/article_vol2.tex
Updated article

%% an abstract and keywords
\Abstract{Data cleaning and -validation are the first steps in any
  data analysis, as the validity of the conclusions from the analysis
  hinges on the quality of the input data. Mistakes in the data can
  arise for any number of reasons, including erroneous codings,
  malfunctioning measurement equipment, inconsistent data generation
  manuals and many more.  Ideally, a human investigator should go
  through each variable in the dataset and look for potential errors
  --- both in input values and coding --- but that process can be very
  time-consuming, expensive and error-prone by itself.

  We describe an \proglang{R} package which implements an extensive
  and customizeable suite of quality assessment tools to be applied to
  a dataset in order to identify potential problems in its
  variables. The results can be presented in an auto-generated,
  non-technical, stand-alone overview document, intended to be perused
  by an investigator with an understanding of the variables in the
  data and the experimental design, but not necessarily knowledge of

  \R{R}. Thereby, \pkg{cleanR} aids the dialogue between data
  analysists and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{cleanR} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{cleanR} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning. \hl{Hvad henvises der til med denne
    sætning?}

@@

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and on data wrangling. While a computer should never
provide an unsupervised decision on what should be done to potential
errors in the dataset, it can be an extremely useful tool in tracking
down and flagging potential erroneous data and in providing
information for humans to easier identify errors based on the context.

@@

While these tools attempt to alleviate the ubiquitous ad hoc approach
to data cleaning they are primarily intended for the data savvy users
and less on the general researcher with a knowledge about the specific
field and context of the available data. The \pkg{cleanR} package
tries to address this by providing a framework that both allows for
extendable, systematic, reproducible data cleaning, and summarizing
findings for researchers from other fields such that they can act as
human experts when tracking down potential errors.

@@

The \pkg{cleanR} approach to data cleaning and -quality assessment is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{cleanR}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} on which data cleaning decisions can be build. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to a inter-field dialogue about the data at
hand. Oftentimes, e.g. distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{cleanR} will help this necessary knowledge sharing.

@@

This manuscript is structured as follows: First, in Section
\ref{sec:usingcleanR}, we introduce the representative of the first
paradigm, namely the \R{clean()} function, which generates data
cleaning overview documents. In the \pkg{cleanR} package, we have
provided a number of default cleaning steps that cover the data
cleaning challenges, we find to be most common. Next, in Section
\hl{3}, we present the interactive mode of \pkg{cleanR}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\hl{4}, we turn to the question of how such \pkg{cleanR} extensions
can be made, such that they are integrable with the \R{clean()}
function and with the other tools available in \pkg{cleanR}.  At last,
in Section 5 \hl{proper ref}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{cleanR} can be used to
solve them.

@@

\section{Creating a data cleaning overview} \label{sec:example1}\label{sec:usingcleanR}

The \R{clean()} function is the primary workhorse of \pkg{cleanR} and
this is the only function needed by the user if the standard battery
of tests are used to output the data cleaning summaries. The data
cleaning output itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data cleaning output document, produced by
calling \R{clean()} on the dataset \R{toyData} available in
\pkg{cleanR}. The first two pages of this data cleaning output are
shown in Figure~\ref{fig:example1}. \R{toyData} is a very
small ($15$ rows and $6$ variables), artificial dataset, whose only purpose is to
illustrate the main capabilities of \pkg{cleanR}. The following
commands load the dataset and produce the cleaning output:

@@

By default, an \proglang{R} markdown file and rendered pdf overview
document is produced, saved to the disc (in the working directory) and
opened for immediate inspection. Turning to Figure~\ref{fig:example1},
we see that such a data cleaning output document consists of two
parts. First, an overview of what was done is presented under the
title \textit{Data cleaning summary}. Secondly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and potentially also a
list of flagged issues. For instance, in the \R{numeric}-type variable
\R{var2} from \R{toyData}, \R{clean()} has identified two values that
are suspected to be miscoded missing values (\R{999} and \R{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.
%We can then return to Part
%1, the data cleaning summary, and inspect what sorts of checks were
%performed on this variable. \hl{hm, delete the last sentence? or use
%  different variable? Only these exact two checks were performed}.

Though the \R{clean()} function is very easy to use, it should not be
mistaken to be inflexible. A large number of function arguments allows
for the cleaning overview document to be molded according to the
user's needs.

The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the cleaning process they influence. In order to understand
this distinction, a glimpse of the inner structure of \R{clean()} is
shown in Figure~\ref{figure:cleanStructure}. In the following we give
examples on how to use these parameters to influence the output of
\R{clean() }.

@@

\smallskip Control input variable and summary\\
\quad \R{useVar} & What variables should be used? & \R{NULL} (corresponding to all variables) \\
\quad \R{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \R{"asIs"} \\
\quad \R{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \R{FALSE} \\
\quad \R{listChecks} & Should an overview of what checks were performed by listed in the \textit{Data cleaning summary}? &  \R{TRUE} \\

\smallskip Control summarize, visualize, and check steps \\
\quad \R{mode} & What steps should be performed for each variable (out of the three possibilities \textit{summarize}, \textit{visualize}, \textit{check})? & \R{c("summarize", "visualize", "check")} \\
\quad \R{labelled\_as} & How should variables of class \R{labelled} be handled (as factors, is missing values or by ignoring labels)? & \R{"factor"} \\
\quad \R{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \R{TRUE} \\
\quad \R{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \R{Inf} \\
\quad \R{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \R{2} \\
\quad \R{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \R{TRUE} \\

\smallskip Control output and post-processing \\
\quad \R{output} & Type of output file to be produced (html, or pdf) & \R{"pdf"} \\
\quad \R{render} & Should the output file be rendered from markdown? & \R{TRUE} \\
\quad \R{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \R{rmarkdown} file be opened? & \R{TRUE} \\
\quad \R{replace} & Overwrite an existing markdown file with the same name? & \R{FALSE} \\

@@

\begin{center}    
\begin{tikzpicture}[node distance = 2cm, auto,thick,scale=0.75, every node/.style={transform shape}]
    % Place nodes
    \node [block] (init) {Get next variable and run pre checks};
    \node [cloud, above of=init] (input) {Input \texttt{data.frame} or \texttt{tibble}};
%    \node [cloud, right of=init] (system) {system};
    \node [decision, right of=init, node distance=4cm] (precheck) {Is variable suitable for inclusion};
    \node [block, right of=precheck, node distance=4cm] (summarize)
    {Run \texttt{summarize()} to produce summary table};
    \node [block, below of=summarize, node distance=3cm] (visualize)
    {Run \texttt{visualize()} to plot variable};
    \node [block, below of=visualize, node distance=2.5cm] (check)
    {Call \texttt{check()} to run error checks};
    \node [decision, below of=check, node distance=2.7cm] (done) {More
      variables?};
    \node [block, right of=done, node distance=3.5cm] (stop) {Write
      \proglang{R} markdown file};
    \node [cloud, below of=stop, node distance=3.5cm] (render) {Render
      markdown and possiby open};
    % Draw edges
    \path [line] (summarize) -- (visualize);
    \path [line] (visualize) -- (check);
    \path [line] (check) -- (done);
    \path [line] (done.south) -- +(0,-10pt) -| node [near start] {yes} (init);
%    \path [line] (identify) -- (evaluate);
%    \path [line] (evaluate) -- (decide);
%    \path [line] (init) -| node [near start] {yes} (precheck);
    \path [line] (init) -- (precheck);
    \path [line] (precheck) -- node [near start] {yes} (summarize);
    \path [line] (precheck) |- node [near start] {no} (done);
    \path [line] (done) -- node [near start] {no} (stop);
%    \path [line] (update) |- (identify);
%    \path [line] (decide) -- node {no}(stop);
    \path [line,dashed] (input) -- (init);
%    \path [line,dashed] (system) -- (init);
%    \path [line,dashed] (system) |- (evaluate);
\end{tikzpicture}
\end{center}

\caption{Overall illustration of the stages undertaken when running
  \R{clean()}. Each variable is checked for eligibility before
  running \R{summarize()}, \R{visualize()}, and \R{check()}, and the
  resulting \proglang{R} markdown file may be rendered and opened.}
\label{figure:cleanStructure}
\end{figure}

@@

For example, to get a summary document in html format that only
contains the variables with potential problems, and with a limit of
maximum 10 printed potential errors for each variable we can write
(output not shown). Also, we can add the argument \R{replace=TRUE} in
order to force \R{clean()} to overwrite any existing files produced by
\R{clean()}.

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html", onlyProblematic=TRUE, maxProbVals=10,
        replace=TRUE)
\end{Sinput}
\end{Schunk}

The final rendering of the generated markdown file is controlled with
the \R{render} and \R{openResult} arguments which both default to
\R{TRUE} \hl{Bliver det ved med at være sandt?}. \R{render} determines
if the \proglang{R} markdown file produced should be rendered using
the \pkg{rmarkdown} package and \R{openResult} decides whether the
output html or pdf file should be opened. The following command
produces an \proglang{R} markdown file containing the data cleaning
process information but the markdown file is not rendered into html
and it is not opened.

@@

\pkg{cleanR} uses three
different types of functions for performing all steps in the above,
namely \code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.
By default, \R{clean()} runs all the summary, visualization, and check
functions currently implemented for each variable type as part of the
algorithm (see
Figure~\ref{figure:cleanStructure}). Table~\ref{table.SVCfunctions}
lists all these functions but we can also use the
\R{allSummaryFunctions()}, \R{allVisualFunctions()}, and
\R{allCheckFunctions()} function in \proglang{R} to print an overview
list. For example the implemented summary functions are:

@@

minMax         Find minimum and maximum        integer, numeric, Date       
               values

quartiles      Compute 1st and 3rd quartiles   integer, numeric

@@

Thus we can see, for example, that the summary table for a
\R{numeric}, \R{integer}, or \R{Date} variable will include
information about the minimum and maximum values, while all variable
classes will produce some kind of information about the central
tendency (i.e., mode or median).

We can control the default summaries and checks for each variable type
through the \R{XXXSummaries} and \R{XXXChecks} arguments where \R{XXX}
represents a variable type, e.g., \R{factorChecks} for factors,
\R{numericChecks} for numeric variables, etc. These arguments accept a
vector of summary or check functions that should be run for a
particular variable type and the default values for each of these
arguments (i.e., the functions from Table~\ref{table.SVCfunctions}
that are applied) can be obtained through the \R{defaultXXXChecks()}
and \R{defaultXXXSummaries()} functions, where \R{XXX} again refers to
the variable type.

@@

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html",  
        characterSummaries = c("variableType", "centralValue"),
        factorSummaries = c("variableType", "centralValue"),
        labelledSummaries = c("variableType", "centralValue"),
        numericSummaries = c("variableType", "centralValue"),
        integerSummaries = c("variableType", "centralValue"),
        logicalSummaries = c("variableType", "centralValue"),
        dateSummaries = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html",  
        allSummaries = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

@@

Now, if we only wanted to apply the identify whitespace function for
factor variables then we would need to set the \R{factorChecks} accordingly

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html",  
        factorChecks = c("identifyWhitespace"))
\end{Sinput}
\end{Schunk}

or we could remove checks for factor altogether by setting the
corresponding argument to \R{NULL} in which case factor variables will
not be checked for any potential errors.

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output="html", factorChecks = NULL)
\end{Sinput}
\end{Schunk}

@@

As indicated in Figure~\ref{figure:cleanStructure} there are two stages
where \R{clean()} applies function(s) to each of the variables: 
\begin{enumerate}
\item In the precheck functions
\item As part of the summarize/visualize/check (SVC) steps
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \R{clean()}, and above we have shown examples of how to
tweak them to modify the data cleaning outputs. However, if for
instance the dataset at hand requires completely different
visualizations then more control is needed, and
Section~\ref{sec:extending} explains how to modify, and expand the
possibilities by producing new functions quite easily.

\section{Using cleanR interactively}
While overview documents are great for presenting and documenting the
data cleaning checks, it may be natural to work more interactively
through the data cleaning process. \pkg{cleanR} also
provides more standard \proglang{R} interactive tools, such as
functions that print results to the console or returns the information
as an object for later use. This section describes how to use the
functions \R{check()}, \R{summarize()} and \R{visualize()} to work
interactively with \pkg{cleanR}.

@@

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \R{check()},
and thus they can also be used here. For instance, if we only want to
check for potentially miscoded missing values, we can use the relevant
\R{XXXChecks} argument (e.g., \R{numericChecks}, \R{factorChecks},
etc. as described in Section~\ref{sec:usingcleanR}) to provide a
vector of the check functions that should be applied. Recall that
Table~\ref{table.SVCfunctions} or the \R{allCheckFunctions()} provides
overviews of the available check functions.
\begin{Schunk}
\begin{Sinput}
> check(toyData$var2, numericChecks = "identifyMissing")
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

An equivalent way to call only a single, specific \R{checkFunction}
such as \R{identifyMissing} (see \ref{table.SVCfunctions} for a list
of check functions) is by using it directly on the variable, e.g.,
\begin{Schunk}
\begin{Sinput}
> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN. 
\end{Soutput}
\end{Schunk}
% $

@@

As we saw with the \R{check()} function, the summary can be modified
by providing the relevant \R{XXXSummaries} arguments. Setting the
\R{numericSummaries} argument we can control the summary output by
providing a vector of functions to run for a particular summary. To
only get the minimum and maximum we set
\R{numericSummaries=c("minMax")}.

\begin{Schunk}
\begin{Sinput}
> summarize(toyData$var2, numericSummaries=c("minMax"))
\end{Sinput}
\begin{Soutput}
     Feature         Result  
[1,] "Min. and max." "1; 999"
\end{Soutput}
\end{Schunk}
% $

>>>>

%% an abstract and keywords
\Abstract{Data cleaning and -validation are the first steps in any
  data analysis, as the validity of the conclusions from the analysis
  hinges on the quality of the input data. Mistakes in the data can
  arise for any number of reasons, including erroneous codings,
  malfunctioning measurement equipment, inconsistent data generation
  manuals and many more.  Ideally, a human investigator should go
  through each variable in the dataset and look for potential errors
  --- both in input values and codings --- but that process can be very
  time-consuming, expensive and error-prone in itself.

  We describe an \proglang{R} package which implements an extensive
  and customizeable suite of quality assessment tools that can be applied to
  a dataset in order to identify potential problems in its
  variables. The results can be presented in an auto-generated,
  non-technical, stand-alone overview document, intended to be perused
  by an investigator with an understanding of the variables in the
  data, but not necessarily knowledge of \R{R}. Thereby, \pkg{cleanR} 
  aids the dialogue between data
  analysts and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{cleanR} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{cleanR} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning, where the data inspections all live and die
  within the \proglang{R} console.

@@

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and on data wrangling. While a computer should never
make unsupervised decisions on what should be done to potential
errors in the dataset, it can still be an extremely useful tool in the 
data cleaning process. Some errors can be tracked down and flagged by a 
computer without further ado, while other types of errors need an analytic
context in order to shine through. But even in the latter case, well-designed
software can aid the process tremendously by giving the human investigator the 
information needed for identifying issues.

@@

While these tools attempt to alleviate the ubiquitous ad hoc approach
to data cleaning, they are primarily intended for the data savvy users
and less so for the general researcher with a knowledge about a specific
field and the context of the available data. The \pkg{cleanR} package
tries to address this by providing a framework that both allows for
extendable, systematic, reproducible data cleaning, and summarizing
findings for researchers from other fields such that they can act as
human experts when tracking down potential errors.

@@

The \pkg{cleanR} approach to data cleaning and -quality assessment is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{cleanR}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} on which data cleaning decisions can be made. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to an inter-field dialogue about the data at
hand. Oftentimes, e.g. distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{cleanR} will help this necessary knowledge sharing.

@@

This manuscript is structured as follows: First, in Section
\ref{sec:usingcleanR}, we introduce the main representative of the first
paradigm, namely the \R{clean()} function, which generates data
cleaning overview documents. In the \pkg{cleanR} package, we have
provided a number of default cleaning steps that cover the data
cleaning challenges, we find to be most common. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{cleanR}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\ref{sec:extending}, we turn to the question of how \pkg{cleanR} extensions
can be made, such that they are integrable with the \R{clean()}
function and with the other tools available in \pkg{cleanR}.  At last,
in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{cleanR} can be used to
solve them.

@@

\section{Creating a data cleaning overview} 
\label{sec:usingcleanR}

The \R{clean()} function is the primary workhorse of \pkg{cleanR} and
this is the only function needed if a standard battery
of tests are used to generate data cleaning summaries. The data
cleaning summary itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data cleaning output document, produced by
calling \R{clean()} on the dataset \R{toyData} available in
\pkg{cleanR}. The first two pages of this data cleaning output are
shown in Figure~\ref{fig:example1}. \R{toyData} is a very
small ($15$ rows of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{cleanR}. The following
commands load the dataset and produce the cleaning output:

@@

By default, an \proglang{R} markdown file and a rendered pdf overview
document is produced, saved to the disc (in the working directory) and
opened for immediate inspection. Turning to Figure~\ref{fig:example1},
we see that such a data cleaning output document consists of two
parts. First, an overview of what was done is presented under the
title \textit{Data cleaning summary}. Secondly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, in the \R{numeric}-type variable
\R{var2} from \R{toyData}, \R{clean()} has identified two values that
are suspected to be miscoded missing values (\R{999} and \R{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.
%We can then return to Part
%1, the data cleaning summary, and inspect what sorts of checks were
%performed on this variable. \hl{hm, delete the last sentence? or use
%  different variable? Only these exact two checks were performed}.

Though the \R{clean()} function is very easy to use, it should not be
mistaken to be inflexible. A large number of function arguments allows
for the cleaning overview document to be molded according to the
user's needs. The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the cleaning process they influence. In order to understand
this distinction, a glimpse of the inner structure of \R{clean()} is
shown in Figure~\ref{figure:cleanStructure}. Below, we present a few
examples on how to use the arguments from Table \ref{table.cleanFormals}
 to influence the output of a \R{clean()} call.

@@

\smallskip Control input variables and summary\\
\quad \R{useVar} & What variables should be used? & \R{NULL} (corresponding to all variables) \\
\quad \R{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \R{"asIs"} \\
\quad \R{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \R{FALSE} \\
\quad \R{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data cleaning summary}? &  \R{TRUE} \\
\quad \R{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \R{c("isKey", "isEmpty")}

\smallskip Control summarize, visualize, and check steps \\
\quad \R{mode} & What steps should be performed for each variable (out of the three possibilities \textit{summarize}, \textit{visualize}, \textit{check})? & \R{c("summarize", "visualize", "check")} (corresponding to all three) \\
\quad \R{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \R{TRUE} \\
\quad \R{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \R{Inf} \\
\quad \R{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \R{2} \\
\quad \R{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \R{TRUE} \\

\smallskip Control output and post-processing \\
\quad \R{output} & Type of output file to be produced (html, or pdf) & \R{"pdf"} \\
\quad \R{render} & Should the output file be rendered from markdown? & \R{TRUE} \\
\quad \R{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \R{rmarkdown} file be opened? & \R{TRUE} \\
\quad \R{replace} & Overwrite an existing files with the same name? & \R{FALSE} \\

@@

\begin{center}    
\begin{tikzpicture}[node distance = 2cm, auto,thick,scale=0.75, every node/.style={transform shape}]
    % Place nodes
    \node [block] (init) {Get next variable and run pre-checks};
    \node [cloud, above of=init] (input) {Input \texttt{data.frame} or \texttt{tibble}};
%    \node [cloud, right of=init] (system) {system};
    \node [decision, right of=init, node distance=4cm] (precheck) {Is variable suitable for inclusion};
    \node [block, right of=precheck, node distance=4cm] (summarize)
    {Run \texttt{summarize()} to produce summary table};
    \node [block, below of=summarize, node distance=3cm] (visualize)
    {Run \texttt{visualize()} to plot variable};
    \node [block, below of=visualize, node distance=2.5cm] (check)
    {Call \texttt{check()} to run error checks};
    \node [decision, below of=check, node distance=2.7cm] (done) {More
      variables?};
    \node [block, right of=done, node distance=3.5cm] (stop) {Write
      \proglang{R} markdown file};
    \node [cloud, below of=stop, node distance=3.5cm] (render) {Render
      markdown and possiby open};
    % Draw edges
    \path [line] (summarize) -- (visualize);
    \path [line] (visualize) -- (check);
    \path [line] (check) -- (done);
    \path [line] (done.south) -- +(0,-10pt) -| node [near start] {yes} (init);
%    \path [line] (identify) -- (evaluate);
%    \path [line] (evaluate) -- (decide);
%    \path [line] (init) -| node [near start] {yes} (precheck);
    \path [line] (init) -- (precheck);
    \path [line] (precheck) -- node [near start] {yes} (summarize);
    \path [line] (precheck) |- node [near start] {no} (done);
    \path [line] (done) -- node [near start] {no} (stop);
%    \path [line] (update) |- (identify);
%    \path [line] (decide) -- node {no}(stop);
    \path [line,dashed] (input) -- (init);
%    \path [line,dashed] (system) -- (init);
%    \path [line,dashed] (system) |- (evaluate);
\end{tikzpicture}
\end{center}

\caption{Schematic illustration of the stages undertaken when running
  \R{clean()}. Each variable is checked for eligibility before
  running \R{summarize()}, \R{visualize()}, and \R{check()}, and the
  resulting \proglang{R} markdown file may be rendered and opened.}
\label{figure:cleanStructure}
\end{figure}

@@

\subsection{Rolling out the arguments}
We begin with an example that is intended as an illustration of how \R{clean()} might
be used in the very first stages of data cleaning, where we are still uncertain
about exactly what is needed and how much times should be allocated to data
cleaning. At this stage, what is really needed, if a very rough idea of the extends of
errors in the dataset. In this scenario, we might wish to obtain a summary document in html format 
that only contains the variables with potential problems, and with a limit of
maximum 10 printed potential errors for each variable we can write
(output not shown). Also, we can add the argument \R{replace=TRUE} in
order to force \R{clean()} to overwrite any existing files produced by
\R{clean()}, as we might run this function a few times in a row before deciding
exactly which cleaning overview we want. Using the \R{toyData} dataset as a guinea pig, we type:

\begin{Schunk}
\begin{Sinput}
> clean(toyData, output = "html", onlyProblematic = TRUE, maxProbVals = 10,
        replace = TRUE)
\end{Sinput}
\end{Schunk}

The final rendering of the generated markdown file is controlled by
the \R{render} and \R{openResult} arguments, which both default to
\R{TRUE} \hl{Bliver det ved med at være sandt?}. \R{render} determines
if the \proglang{R} markdown file produced should be rendered using
the \pkg{rmarkdown} package and \R{openResult} decides whether the
output html or pdf file should be opened. The following command
produces an \proglang{R} markdown file containing the data cleaning
summary information, but without rendering nor opening the markdown file:

@@

We will now move on to discuss how not only the structure of the cleaning summary is manipulated, but also its very contents. This is done through the summarize-visualize-check (SVC) step, as illustrated in Figure \ref{figure:cleanStructure}. 
\pkg{cleanR} uses three different types of functions for performing these steps,
namely \code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.
By default, \R{clean()} runs selected summary, visualization and check functions on each variable in the dataset, and the exact choice of these functions depends on the classes of the variables. For instance, though detection of outlier values might be interesting for numerical variables, it holds little meaning for factor variables, and therefore, numerical and factor variables need different checks. Table~\ref{table.SVCfunctions}
lists all available SVC functions, but we can also use the
\R{allSummaryFunctions()}, \R{allVisualFunctions()}, and
\R{allCheckFunctions()} functions in \pkg{cleanR} to print overview
lists in \proglang{R}. For example, the implemented \R{summaryFunction}s are:

@@

minMax         Find minimum and maximum        Date, integer, numeric        
               values

quartiles      Compute 1st and 3rd quartiles   Date, integer, numeric

@@

Thus we can see, for example, that for \R{numeric}, \R{integer}, or \R{Date} 
variables, \pkg{cleanR} provides functions for adding summary information about 
the minimum and maximum values, while all seven variable
classes dealt with in \pkg{cleanR} have functions for central
tendency summaries (i.e., mode or median).

We can control what summaries and checks are applied for each variable type
through the \R{XXXSummaries} and \R{XXXChecks} arguments, where \R{XXX}
represents a variable type, e.g., \R{factorChecks} for factors,
\R{numericChecks} for numeric variables, etc. These arguments accept a
vector of summary- or check function names that should be run for a
particular variable type. The default values for each of these
arguments can be obtained through the \R{defaultXXXChecks()}
and \R{defaultXXXSummaries()} functions, where \R{XXX} again refers to
the variable type.

@@

\begin{Schunk}
\begin{Sinput}
> clean(toyData, characterSummaries = c("variableType", "centralValue"),
        factorSummaries = c("variableType", "centralValue"),
        labelledSummaries = c("variableType", "centralValue"),
        numericSummaries = c("variableType", "centralValue"),
        integerSummaries = c("variableType", "centralValue"),
        logicalSummaries = c("variableType", "centralValue"),
        dateSummaries = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> clean(toyData, allSummaries = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

@@

Now, if we only wanted to apply the identify whitespace function for
factor variables, then we would need to set the \R{factorChecks} accordingly

\begin{Schunk}
\begin{Sinput}
> clean(toyData, factorChecks = c("identifyWhitespace"))
\end{Sinput}
\end{Schunk}

or we could remove checks for factors altogether by setting the
corresponding argument to \R{NULL}, in which case factor variables will
not be checked for any potential errors:

\begin{Schunk}
\begin{Sinput}
> clean(toyData, factorChecks = NULL)
\end{Sinput}
\end{Schunk}

As with \R{summaryFunction}s, a complete list of available \R{checkFunction}s is obtained by calling 
\R{allCheckFunctions()}. Note however, that \R{checkFunction}s have a usage beyond the 
\R{XXXChecks} arguments, namely in the \textit{pre-check} stage. In this stage, it is determined 
whether or not each variable is suitable for the SVC-steps. The functions used in the pre-check stage
should be \R{checkFunction}s that are applicable to all variable classes. The default pre-checks, the functions
\R{isKey} and \R{isEmpty}, check whether a variable has unique values for all observations or only a single value for 
all observations, respectively. If one of these statements are true, the variable will not be subjected to the SVC steps. 
We can allow empty variables to move on to the SVC step by only checking for keys in the pre-check step:

\begin{Schunk}
\begin{Sinput}
> clean(toyData, preChecks = "isKey")
\end{Sinput}
\end{Schunk}

Note that the data visualizations in the cleaning summary are also controllable, though only a single function can be provided for all variable types. If for instance, we wish to change the visualizations from the default \pkg{ggplot2} style histograms and barplots to base \proglang{R} histograms and barplots, we type
\begin{Schunk}
\begin{Sinput}
> clean(toyData, allVisuals = "basicVisual")
\end{Sinput}
\end{Schunk}

@@

In summary, and as indicated in Figure~\ref{figure:cleanStructure}, there are two stages
where \R{clean()} applies functions to each of the variables: 
\begin{enumerate}
\item In the pre-check stage
\item As part of the summarize/visualize/check (SVC) steps
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \R{clean()}, and above we have shown examples of how to
tweak them to modify the data cleaning outputs. However, if for
instance the dataset at hand requires new, additional checks, then more control is needed, and
Section~\ref{sec:extending} explains how to modify and expand the
possibilities by producing new summary-, visual- and check functions.

\section{Using cleanR interactively}
\label{sec:interactiveCleanR}

While overview documents are great for presenting and documenting the
data cleaning checks, it may be useful to be able to work more interactively
through the data cleaning process as well. Aside from the \R{clean()} function 
presented above, \pkg{cleanR} also provides more standard \proglang{R} interactive tools, such as
functions that print results to the console or returns the information
as an object for later use. This section describes how to use the
functions \R{check()}, \R{summarize()} and \R{visualize()} to work
interactively with \pkg{cleanR}.

@@

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \R{check()},
and thus they can also be used here. For instance, if we only want to
check for potentially miscoded missing values, we can use the relevant
\R{XXXChecks} argument (e.g., \R{numericChecks}, \R{factorChecks},
etc. as described in Section~\ref{sec:usingcleanR}) to provide a
vector of the check functions that should be applied. Recall that
Table~\ref{table.SVCfunctions} or an \R{allCheckFunctions()} call provide
overviews of the available check functions. 
Moving forward, we limit the numeric checks to only identify miscoded 
missing values:

\begin{Schunk}
\begin{Sinput}
> check(toyData$var2, numericChecks = "identifyMissing")
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

An equivalent way to call only a single, specific \R{checkFunction},
such as \R{identifyMissing}, is by using it directly on the variable, e.g.,
\begin{Schunk}
\begin{Sinput}
> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN. 
\end{Soutput}
\end{Schunk}
% $

@@

As we saw with the \R{check()} function, the summary can be modified
by providing the relevant \R{XXXSummaries} arguments. Setting the
\R{numericSummaries} argument, we can control the summary output by
providing a vector of function names to run for a particular summary. To
only get the median, the minimum and the maximum we set
\R{numericSummaries=c("centralValue", "minMax")}:

\begin{Schunk}
\begin{Sinput}
> summarize(toyData$var2, numericSummaries = c("centralValue", "minMax"))
\end{Sinput}
\begin{Soutput}
     Feature         Result  
[1,] "Median"        "4.5"   
[2,] "Min. and max." "1; 999"
\end{Soutput}
\end{Schunk}
% $
